---
title: "nutrient-mvco-eml-assembly"
author: "Kate Morkeski"
date: "2/3/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load required packages 

# two of the required packages are instinalled from GitHub
# uncomment these lines to update as needed
# library(remotes)
# remotes::install_github("EDIorg/EMLassemblyline")
# remotes::install_github("WHOIGit/ediutilities")

library(EMLassemblyline)
library(ediutilities)
library(here)
library(tidyverse)
library(readxl)
library(lubridate)
library(devtools)
library(EML)
library(maps)
library(xml2)
library(geosphere)
library(httr)
library(compareDF)
library(dplyr)

#set path to root of project
here("nes-lter-nutrient")

```

## Read in provided data
```{r}

nutrients_input <- read_excel('MVCO_nut_copy20240604.xls') 

```
## Update column names

```{r}

# remove quotation marks and spaces
names(nutrients_input)<- gsub("'", "", names(nutrients_input))
names(nutrients_input)<- gsub(" ", "", names(nutrients_input))
names(nutrients_input)<- gsub("-", "", names(nutrients_input))
names(nutrients_input)<- gsub(" + ", "", names(nutrients_input))

# combine date and time
nutrients_input$Start_Date <- as.character(nutrients_input$Start_Date) 
nutrients_input$Start_Time_UTC <- as.character(nutrients_input$Start_Time_UTC)
nutrients_input$Start_Time_UTC <-gsub("1899-12-31 ", "", nutrients_input$Start_Time_UTC)
nutrients_input$datetime <- paste(nutrients_input$Start_Date, nutrients_input$Start_Time_UTC)
nutrients_input$datetime <- as.POSIXct(nutrients_input$datetime, tz = "GMT", format="%Y-%m-%d %H:%M:%OS")

# rename nutrient columns
nutrients_input <-  nutrients_input %>% 
  rename(nitrate_a = `Nut_a_uMNO2+NO3`) %>%  # nitrate is really nitrate + nitrite #keep simple for wrangling then rename
  rename(nitrate_b = `Nut_b_uMNO2+NO3`) %>%
  rename(nitrate_c = `Nut_c_uMNO2+NO3`) %>%
  rename(ammonium_a = `Nut_a_uMNH4+`) %>%
  rename(ammonium_b = `Nut_b_uMNH4+`) %>%
  rename(ammonium_c = `Nut_c_uMNH4+`) %>%
  rename(silicate_a = `Nut_a_uMSiO2`) %>%
  rename(silicate_b = `Nut_b_uMSiO2`) %>%
  rename(silicate_c = `Nut_c_uMSiO2`) %>%
  rename(phosphate_a = `Nut_a_uMPO43`) %>%
  rename(phosphate_b = `Nut_b_uMPO43`) %>%
  rename(phosphate_c = `Nut_c_uMPO43`) %>%
  select(-Start_Date, -Start_Time_UTC)  %>%
  relocate(datetime, .before = Latitude)

names(nutrients_input)<- tolower(names(nutrients_input))

# round lat/long and nutrient concentration columns
nutrients_input <- nutrients_input  %>% 
  mutate(latitude = round(latitude, 4)) %>% 
  mutate(longitude = round(longitude, 4)) %>% 
  mutate(nitrate_a = round(nitrate_a, 3)) %>% 
  mutate(nitrate_b = round(nitrate_b, 3)) %>% 
  mutate(nitrate_c = round(nitrate_c, 3)) %>% 
  mutate(ammonium_a = round(ammonium_a, 3)) %>% 
  mutate(ammonium_b = round(ammonium_b, 3)) %>% 
  mutate(ammonium_c = round(ammonium_c, 3)) %>% 
  mutate(phosphate_a = round(phosphate_a, 3)) %>% 
  mutate(phosphate_b = round(phosphate_b, 3)) %>% 
  mutate(phosphate_c = round(phosphate_c, 3)) %>% 
  mutate(silicate_a = round(silicate_a, 3)) %>% 
  mutate(silicate_b = round(silicate_b, 3)) %>% 
  mutate(silicate_c = round(silicate_c, 3))  

# ensure rows are in time order
nutrients_input <- nutrients_input[order(nutrients_input$datetime),]

# create dataframe for comparison to package version 1 data
# remove event_number_niskin field from this dataframe only
nutr_input_compare <- nutrients_input %>%
 select(-event_number_niskin)

```

## Pivot current data table to handle replicates

```{r}

nitrate <- nutrients_input %>% 
 select(-ammonium_a, -ammonium_b, -ammonium_c) %>%
 select(-silicate_a, -silicate_b, -silicate_c) %>%
 select(-phosphate_a, -phosphate_b, -phosphate_c)  %>% 
 pivot_longer(cols = c(nitrate_a, nitrate_b, nitrate_c), names_to = "replicate",  values_to = "nitrate")  
nitrate$replicate <- gsub("nitrate_", "", nitrate$replicate)
  
ammonium <- nutrients_input %>% 
  select(-nitrate_a, -nitrate_b, -nitrate_c) %>%
  select(-silicate_a, -silicate_b, -silicate_c) %>%
  select(-phosphate_a, -phosphate_b, -phosphate_c) %>% 
  pivot_longer(cols = c(ammonium_a, ammonium_b, ammonium_c), names_to = "replicate",  values_to = "ammonium") 
ammonium$replicate <- gsub("ammonium_", "", ammonium$replicate)
  
silicate <- nutrients_input %>%  
  select(-nitrate_a, -nitrate_b, -nitrate_c) %>%
  select(-ammonium_a, -ammonium_b, -ammonium_c) %>%
  select(-phosphate_a, -phosphate_b, -phosphate_c) %>%
  pivot_longer(cols = c(silicate_a, silicate_b, silicate_c), names_to = "replicate",  values_to = "silicate") 
silicate$replicate <- gsub("silicate_", "", silicate$replicate)
  
phosphate <- nutrients_input %>% 
  select(-nitrate_a, -nitrate_b, -nitrate_c) %>%
  select(-ammonium_a, -ammonium_b, -ammonium_c) %>%
  select(-silicate_a, -silicate_b, -silicate_c) %>%
  pivot_longer(cols = c(phosphate_a, phosphate_b, phosphate_c), names_to = "replicate",  values_to = "phosphate")
phosphate$replicate <- gsub("phosphate_", "", phosphate$replicate)

nutrients_long <- full_join(nitrate, ammonium, by = c("event_number", "event_number_niskin", "datetime", "latitude", "longitude", "depth", "replicate"))
nutrients_long <- full_join(nutrients_long, silicate, by = c("event_number", "event_number_niskin", "datetime", "latitude", "longitude", "depth", "replicate"))
nutrients_long <- full_join(nutrients_long, phosphate, by = c("event_number", "event_number_niskin", "datetime", "latitude", "longitude", "depth", "replicate"))

```

# QA: transformation check

```{r}

# check that number of samples and sums of values in transformed data match input data

nutrient <- c("nitrate", "ammonium", "phosphate", "silicate")
check <- data.frame(nutrient)
check$input_n <- NA_integer_
check$input_sum <- NA_integer_
check$long_n <- NA_integer_
check$long_sum <- NA_integer_


# count non-NA values in input data and sum all values of each nutrient in input data
for (i in 1:nrow(check)){
  
  nutrient <- check$nutrient[i]
 
  a <- select(nutrients_input, paste0(nutrient,"_a"))
  b <- select(nutrients_input, paste0(nutrient,"_b"))
  c <- select(nutrients_input, paste0(nutrient,"_c"))
  
  check$input_n[i] <- sum(!is.na(a)) + sum(!is.na(b)) + sum(!is.na(c))
  check$input_sum[i] <- sum(a, na.rm = TRUE) + sum(b, na.rm = TRUE) + sum(c, na.rm = TRUE)
  
}

# count non-NA values in transformed data and sum all values of each nutrient in transformed data
for (i in 1:nrow(check)){
  
  nutrient <- check$nutrient[i]
 
  check$long_n[i] <- sum(!is.na(select(nutrients_long, nutrient)))
  check$long_sum[i] <- sum(select(nutrients_long, nutrient), na.rm = TRUE)
  
}

check$diff_n <- check$input_n - check$long_n 
check$diff_sum <- check$input_sum - check$long_sum

## visually inspect check dataframe for any non-zero diff_n and diff_sum

```

# Clean up long dataframe

```{r}

# remove rows containing no nutrient data

nutrients <- nutrients_long %>%
  filter(!is.na(nitrate | ammonium | silicate | phosphate))

# confirm results # can also check against dataframe "check"

```
# Add quality flag for phosphate samples

```{r}

# add quality flag for phosphate samples to reflect concern of possible contamination
# for events MVCO_304 to MVCO_307 flag = 3 = questionable
# all else flag = 1 = good

nutrients <- nutrients %>%
   mutate(iode_quality_flag_phosphate = case_when(event_number == "MVCO_304" ~ 3,
                                                  event_number == "MVCO_305" ~ 3,
                                                  event_number == "MVCO_306" ~ 3,
                                                  event_number == "MVCO_307" ~ 3,
                                                  TRUE ~ 1))

```

# Add Tioga cruise number

```{r}

df <- read_excel('List of MVCO cruises for NES-LTER project.xlsx', sheet = "catalog", skip = 6)
Tioga_ID <- df %>% select(1, 3, 4, 5)

nutrients <- left_join(nutrients, Tioga_ID, by = "event_number")

nutrients <- nutrients %>%
  select(-date)  %>%
  relocate(ship, .before = datetime) %>%
  relocate(cruise_ID, .before = datetime)
 
```

# Merge MVCO samples from transect cruises

```{r}
base_api_url <- 'https://nes-lter-data.whoi.edu/api'

cruise_list <- api_list_cruises()

all_cruises <- NA
print(cruise_list)
for(cruise_id in cruise_list) {
  cruise_id <- str_to_lower(cruise_id)
  nut_url <- glue::glue('{base_api_url}/nut/{cruise_id}.csv')
  print(nut_url)

  cruise_nut <- tryCatch(read_csv(nut_url), 
                 error=function(cond) 
                     { message(cond)
                       return(NULL) })

  if(nrow(cruise_nut) > 0 && !is.null(cruise_nut)) {
    if(length(all_cruises) == 1 && is.na(all_cruises)) {
      all_cruises <- cruise_nut
    } else {
      all_cruises <- plyr::rbind.fill(all_cruises, cruise_nut)
    }
  }
}

# Transect MVCO ranges lat =  41.3125 - 41.3269, lon = -70.5629 to -70.5857
# MVCO_nut_copy ranges lat = 41.1364 - 41.339, lon = -70.4151 to -70.6839
lat_min <- 41.13
lat_max <- 41.33
lon_min <- -70.68
lon_max <- -70.41

# Filter rows within the specified range of latitude and longitude
filtered_nut <- all_cruises %>%
  filter(latitude >= lat_min & latitude <= lat_max,
         longitude >= lon_min & longitude <= lon_max)

filtered_nut <- filtered_nut %>%
  rename(nitrate = nitrate_nitrite,
         datetime = date)

# Add ship, event_number, event_number_niskin, and cruise_ID columns and populate
filtered_nut <- filtered_nut %>%
  mutate(ship = case_when(
    grepl("^AR", cruise) ~ "Armstrong",
    grepl("^AT", cruise) ~ "Atlantis",
    grepl("^EN", cruise) ~ "Endeavor",
    grepl("^HRS", cruise) ~ "Sharp",
    TRUE ~ ""  # Set to empty string for unmatched cases
      ),
    event_number = paste(cruise, cast, sep = "_"),  
    event_number_niskin = paste(cruise, cast, niskin, sep = "_"), 
    cruise_ID = cruise
  )

# Merge transect nut data with nutrients
common_columns <- intersect(names(nutrients), names(filtered_nut))
filtered_nut_aligned <- filtered_nut %>% select(all_of(common_columns))
missing_columns <- setdiff(names(nutrients), names(filtered_nut_aligned))
filtered_nut_aligned[missing_columns] <- NA
filtered_nut_aligned <- filtered_nut_aligned %>% select(names(nutrients))
nutrients_updated <- bind_rows(nutrients, filtered_nut_aligned)

# Round lat/long and nutrient concentrations
nutrients_updated <- nutrients_updated  %>% 
  mutate(latitude = round(latitude, 4)) %>% 
  mutate(longitude = round(longitude, 4))  %>% 
  mutate(nitrate = round(nitrate, 3))  %>% 
  mutate(ammonium = round(ammonium, 3))  %>% 
  mutate(silicate = round(silicate, 3))  %>% 
  mutate(phosphate = round(phosphate, 3))

# Set iode_quality_flag_phosphate to 2 where it is NA
nutrients_updated <- nutrients_updated %>%
  mutate(iode_quality_flag_phosphate = ifelse(is.na(iode_quality_flag_phosphate), 2, iode_quality_flag_phosphate))

nutrients <- nutrients_updated[order(nutrients_updated$datetime), ]

```

## QA: Determine if any nutrient values exceed expectations

According to a global range check: nitrate less than 30 umol/l and ammonium less than 5 umol/l based on [Rees et al. 2006](https://doi.org/10.1016/j.dsr2.2006.05.008), phosphate less than 3 (no great reference but this appears to be upper for Atlantic), silicate less than 60 (Elements of Physical Oceanography chapter on marine silica cycle, for deep Atlantic, this is probably too high).

```{r}

summary(nutrients) # visual min max for all columns
# visually check the printed nutrient data for maxima exceeding expected thresholds:
# nitrate > 30 
# ammonium > 5
# phosphate > 3
# silicate > 60

# filter for values exceeding the thresholds
high_nitrate <- nutrients %>%  filter(nitrate > 30)
high_ammonium <- nutrients %>% filter(ammonium > 5)
high_phosphate <- nutrients %>% filter(phosphate > 3)
high_silicate <- nutrients %>% filter(silicate > 60)

# check console output for summary and alerts
print(paste(nrow(high_nitrate), "nitrate observations exceed max" ))
print(paste(nrow(high_ammonium), "ammonium observations exceed max" ))
print(paste(nrow(high_phosphate), "phosphate observations exceed max" ))
print(paste(nrow(high_silicate), "silicate observations exceed max" ))
 
# investigate high ammonium values
hist(high_ammonium$ammonium)

# could rbind all high nutrient data frames for better reproducibility # this way is simpler in this instance
write.csv(high_ammonium, "high_ammonium.csv")
# only one of these values is new to version 3

```
## QA: Plot nutrients as a check

Make boxplots based on all nutrient values

```{r}

# define the nutrient columns 
nut_cols <- c("nitrate", "phosphate", "ammonium", "silicate")
# convert to long for ease of plotting
nutrients_longer <- nutrients %>%
    gather(nutrient, value, all_of(nut_cols), factor_key = TRUE)

# boxplot of the 4 nutrients in single panel
ggplot(nutrients_longer, aes(x=nutrient, y=value))+geom_boxplot(outlier.size = 0.5)+theme_classic()

ggsave("4-nutrients-single-panel.png")

# boxplots of individual nutrients
for (i in 1:length(nut_cols)) {
  nut_subset <- nutrients_longer %>% filter(nutrient == nut_cols[i])

  # ggplot where x = cast, y = value and the lineplots are grouped by cruise
  p <- ggplot(data = nut_subset, aes(x = nutrient, y = value)) +
    geom_boxplot(outlier.size = 0.5) +
    ylab(paste0(nut_cols[i], " concentration (µmol/L)")) +
    theme_classic()
  ggsave(paste0("nutrients-individual-", nut_cols[i], ".png"), plot = p)
}

ggplot(nutrients, aes(x=as.factor(iode_quality_flag_phosphate), y=phosphate)) + 
   geom_boxplot() + xlab('iode_quality_flag')

ggsave("nutrients-phosphate.png")

```

# QA: Nutrient Outlier Check

# Check if there are severe differences in nutrient values across replicates. Plot these differences to perform a visual check.

```{r}

# calculate the difference between the replicates across all nutrients
nut_check <- nutrients %>%
  group_by(event_number_niskin) %>%
    mutate(nitrate_diff = abs(nitrate - lead(nitrate)),
          ammonium_diff = abs(ammonium - lead(ammonium)),
          phosphate_diff = abs(phosphate - lead(phosphate)),
          silicate_diff = abs(silicate - lead(silicate))) 

# define the nutrient outlier columns to gather on
nut_diff_cols <- c("nitrate_diff", "phosphate_diff", "ammonium_diff", "silicate_diff")
# convert to long for ease of plotting
nut_check_long <- nut_check %>%
  select(nitrate_diff, ammonium_diff, phosphate_diff, silicate_diff) %>%
  filter(nitrate_diff != 0) %>%
  gather(nutrient_diffs, value, all_of(nut_diff_cols))

# plot the differences and look for outliers
ggplot(data = nut_check_long, aes(x = nutrient_diffs, y = value)) +
    geom_boxplot(outlier.size = 0.5) +
    ylab(paste0("Concentration (µmol/L)")) +
    theme_classic()

ggsave("nutrients-differences.png")

```

## QA map check of sampling locations

```{r}

# plot location of all samples
map_locs(nutrients, colorvar = "depth", region = "transect")
# map_locs(nutrients_long, colorvar = "datetime", region = "transect")

# investigate distribution of low-latitude samples
hist(nutrients$latitude, plot = FALSE)

# filter samples into northern and southern sections using cutoff of 41.30 deg N suggested by histogram data
southern_samples <- nutrients %>% filter(latitude < 41.30 )
northern_samples <- nutrients %>% filter(latitude > 41.30 )
# southern samples are from 2007-2009 so already in package version 1

# plot sections 
map_locs(southern_samples, colorvar = "depth", region = "transect")
map_locs(southern_samples, colorvar = "ammonium", region = "transect")
#map_locs(northern_samples, colorvar = "depth", region = "transect")

# plot high ammonium samples
map_locs(high_ammonium, colorvar = "ammonium", region = "transect")

new_data <- nutrients %>% filter(datetime > "2017-05-30 20:00:00")
map_locs(new_data, colorvar = "ammonium", region = "transect")

new_old_data <- nutrients %>% filter(datetime < "2006-04-19 21:10:00")
map_locs(new_old_data, colorvar = "ammonium", region = "transect")

version1  <- nutrients %>% filter("2006-04-19 21:10:00" < datetime & datetime < "2017-05-30 20:00:00")
map_locs(version1, colorvar = "ammonium", region = "transect")

hist(nutrients$longitude, plot = FALSE)
east_west <- nutrients %>% filter(-70.53 < longitude | longitude < -70.59)
map_locs(east_west, colorvar = "ammonium", region = "transect")

```
# Format data for package

```{r}

# rename nitrate column to nitrate+nitrite
nutrients <- rename(nutrients, nitrate_nitrite = nitrate) 

nutrients <- nutrients %>%
  mutate(
    ship = ifelse(is.na(ship), "", ship),
    cruise_ID = ifelse(is.na(cruise_ID) | cruise_ID == "NA", "", cruise_ID)
  )

```

## Read in previous version MVCO nutrient package to compare to new data set

```{r}

url <- ('https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-nes.1.3&entityid=077bb6d32f84c773852b678f1fe52e57')

httr::GET(url, write_disk(tf <- tempfile(fileext = ".csv")))
tf

nutr_pubd <- read_csv(tf, TRUE, show_col_types = FALSE) 

comparison <- compare_df(nutrients, nutr_pubd, c("event_number", "depth"))

create_output_table(comparison, output_type='xlsx', file_name='version_comparison.xlsx')

```

## Create package data file

```{r}
  
# write csv file for package
write.csv(nutrients, "nes-lter-nutrient-mvco.csv", na = "NaN", row.names = FALSE)

```

## EML Assembly

Read data table

```{r}

data_table <- read.csv(here('nes-lter-nutrient-mvco.csv'))

# identify date column for ediutilities package to generate temporal coverage
data_table$date = ymd_hms(data_table$datetime)

```

Read the Excel metadata template and generate text templates used by
EMLassemblyline

```{r}  

excel_to_template(here('nutrient-mvco-info'), 'nes-lter-nutrient-mvco', rights='CCBY')  

```
Generate the package and insert the parent project node into the resulting EML

```{r}

pkg_id <- 'knb-lter-nes.1.4'  
make_eml(here(),
         dataset.title="Dissolved inorganic nutrients from the Martha's Vineyard Coastal Observatory (MVCO), including 4 macro-nutrients from water column bottle samples, ongoing since 2003 (NES-LTER since 2017)",
         data.table='nes-lter-nutrient-mvco.csv',
         data.table.description='Dissolved inorganic nutrients from water column bottle samples taken near the Marthas Vineyard Coastal Observatory',
         data.table.name = 'nes-lter-nutrient-mvco',
         temporal.coverage = temporal_coverage(data_table$date),
         geographic.description = "Marthas Vineyard Coastal Observatory",
         geographic.coordinates = geographic_coordinates(data_table$latitude, data_table$longitude),
         maintenance.description = "ongoing",
         user.id = "NES",
         user.domain = "LTER",
         package.id = pkg_id)
project_insert(pkg_id)

```

